Order statistics and numerical optimization: the case of a single observation

### 2020-03-04 inst/skating_analysis.Rmd
more extensive changes reflecting ordertable_to_ranktable()

This script creates and analyses hyper2 object "skating" which is a
likelihood function for the strengths of the competitors in the
Ladies' Figure Skating at the 2002 Winter Olympics.  Dataframe
`skating_table` is copied from Lock and Lock (see skating.Rd for
more details).

```{r}
library("hyper2",quietly=TRUE)
set.seed(0)
data("skating")
skating_table
```

Object `skating_table` is structured so that each competitor is a
row, and each judge is a column.  Function `rank_likelihood()`
considers each row to be a race [or a judge], so we need to take a
transpose.  The following R idiom is plausible but incorrect:

@@

The above is incorrect because we want the ranks, and we have the
order.  Correct analysis follows.

```{r}
skating_rank <- apply(skating_table,2,order)
head(skating_rank)
```

Column 1 of `skating_table` and column 1 of `skating_rank` are
confusingly simular.  So consider column 2 instead, corresponding to
judge number 2.  If `a=skating_table[,2]` and
`b=skating_rank[,2]` we have `a=c(4,1,3,2, ...)` and
`b=c(2,4,3,1, ...)`.  Thus `a` means that hughes was judged to
be fourth best, slutskaya was judged to be the best, kwan third best,
cohen second best, and so on.  Conversely, `b` means that the best
competitor was number two [slutskaya], the second best number was
number four [cohen], the third best was number three [kwan], the
fourth best number 1 [hughes], and so on.  _Note that this means the
rows of `skating_rank` cannot be named_.

```{r}
skating <- rank_likelihood(t(skating_rank))
pnames(skating) <- rownames(skating_table)
head(skating)
```

@@

In the above, each row is a competitor and each column corresponds to
a ranking.  Thus the first column corresponds to "first place" and
	shows that five judges ranked Hughes first and four judges ranked
Slutskaya first.  The second column corresponds to "second place" and
shows that one judge gave Hughes second place, two judges placed
Slutskaya second, five gave Kawn second, and one gave Cohen second.
The formal ordering is, I think, to rank the competitors by number of
first places, then second, then third , and so on.  Ties at level $n$
are broken by considering $n+1$.

@@

small <- 1e-6  # need a sensible start value satisfying the constraints
start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))

@@

for(i in seq_len(nrow(problem_constraints))){
jj <- maxp(skating, startp=start, give=TRUE,fcm=problem_constraints[i,], fcv=0)
   fullout[[i]] <- jj
   out[i] <- jj$value
}
out
```

@@

```{r,unconstrainedlikes}
mgv <- maxp(skating,give=TRUE)$value
mgv  - out
mgv  - max(out)
```

>>>>

```{r, echo=T, results = 'hide', message= FALSE}
library("hyper2",quietly=TRUE)
```

This document analyses strengths of the competitors in the Ladies'
Figure Skating at the 2002 Winter Olympics.  In the package, dataframe
`skating_table` is copied from Lock and Lock (see skating.Rd for more
details).

```{r,showskating}
set.seed(0)
data("skating")
skating_table
```

Object `skating_table` is an **order table**.  It is structured so
that each competitor is a row, and each judge is a column.  Function
`rank_likelihood()` considers each row to be a race [or a judge], so
we need to take a transpose.  The following R idiom is plausible but
incorrect:

@@

The above is incorrect because we are applying a rank technique to an
order table.  Correct analysis follows; firstly we coerce to a
**rank** table:

```{r}
(skating_order <- ordertable_to_ranktable(skating_table))
```

In the above, the column names should be read "came first" (`c1`),
"came second" (`c2`) and so on to "came 23rd" (`c23`).  From the first
column, we see that Hughes came first 5 times and Slutskya came first
four times, and from the last column we see that all judges except
`J3` awarded Luca last place.  File `man/rrank.Rd` has a detailed
discussion of the differences between rank and order.

We can coerce from an order table to a likelihood function:

```{r}
head(ordertable2supp(skating_order))
```

@@

In the above, each row is a competitor and each column corresponds to
a ranking.  Thus the first column corresponds to "first place" and
shows that five judges ranked Hughes first and four judges ranked
Slutskaya first.  The second column corresponds to "second place" and
shows that one judge gave Hughes second place, two judges placed
Slutskaya second, five gave Kawn second, and one gave Cohen second.

The formal ordering used in competition is, according to Lock and
Lock, given by the median ordinal:

```{r,medianordinal}
apply(skating_table,1,median)
```

but with ties broken using a complicated hierarchy, the first of which
is the "size of the majority":

```{r,sizeofthemajority}
rowSums(sweep(skating_table,1,seq_len(23))>=0)
```

which would suggest that Slutskaya beats Kwan (5-4), Butyrskaya draws
with Robinson (7-7), and so on.  The rows of `skating_table` are in
the order given by this system.

@@

small <- 1e-4  # need a sensible start value satisfying the constraints
start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))

@@

for(i in seq_len(nrow(problem_constraints))){
  jj <- maxp(skating, startp=start, give=TRUE,fcm=problem_constraints[i,], fcv=0,n=1,SMALL=1e-5)
   fullout[[i]] <- jj
   out[i] <- jj$value
}
out
```

@@

```{r,unconstrainedlikes}
mgv <- maxp(skating, give=TRUE, n=2)$value
mgv  - out
mgv  - max(out)
```
